---
title: "LPS Switch"
date: today
embed-resources: true
execute:
    echo: false
    warning: false
editor_options: 
  chunk_output_type: console
---
```{r}

library(tidyverse)
library(DESeq2)          # Framework for RNA-seq differential expression
library(patchwork)
library(compareGroups)
library(kableExtra)
library(EnhancedVolcano)

# Read the raw gene count data from a tab-separated value (TSV) file.
counts <- read.table("~/Dropbox/SyncBriefcase/LAB/UK/APOE_switch/Georgia_bulk/nf-core_results/salmon_counts/salmon.merged.gene_counts.tsv", 
                       header = TRUE, sep = "\t") |>  # Specify that the file has a header and is tab-delimited.
  # Remove any duplicate rows based on the 'gene_name' column, keeping the first occurrence.
  distinct(gene_name, .keep_all = T) |> 
  # Set the 'gene_name' column as the row names of the data frame. DESeq2 requires gene identifiers as row names.
  column_to_rownames("gene_name") |> 
  # Remove the 'gene_id' and 'M2' columns, which are not needed for this analysis.
  select(-gene_id, -M2) |> 
  # Convert all remaining columns (which are sample counts) to integer format, as required by DESeq2.
  mutate(across(everything(), as.integer))


# Read the sample metadata from an Excel file into a data frame called 'md0'.
md0 <- readxl::read_xlsx("~/Dropbox/SyncBriefcase/LAB/UK/APOE_switch/Georgia_bulk/20250905_Bulk seq IDs.xlsx")

# Process the metadata.
md <- md0 |>
    # Set the 'altID' column as the row names for the metadata frame 'md'. 
    # These row names should correspond to the sample names (column names) in the 'counts' data frame.
    column_to_rownames("altID") 
    
# Reorder the rows of the metadata data frame ('md') to match the column order of the counts data frame ('counts').
# This is a CRITICAL step to ensure that the sample information in 'md' correctly aligns with the sample data in 'counts'.
md <- md[colnames(counts),] 

# identical(rownames(md),names(counts))


######################################################
### cytokine data ####################################
######################################################

ck0 <- read_csv("~/Dropbox/SyncBriefcase/LAB/UK/APOE_switch/Georgia_bulk/Raw Data Johnson_MacLean Cytokine Results 2023 0612.csv")

ck <- separate_wider_delim(ck0, cols = "Name", names = c("sample","tissue","prepost","experiment"), 
                           delim = " ",
                           too_few = "align_start") |> 
    mutate(experiment = ifelse(prepost == "LPS", "LPS", experiment ),
           prepost = ifelse (prepost == "LPS", "Brain", prepost)) |> 
    filter( experiment == "LPS") |> 
    left_join(md, by = c ("sample" = "mouseID")) |> 
    filter(!is.na(genotype)) |> 
    mutate(across(IFN:IL6, \(x) parse_number(x)),
           prepost = factor (prepost, levels = c("(Pre)","(Post)","Brain")))
    
ck_wider <- ck |> 
    select(-tissue) |> 
    pivot_wider(names_from = prepost, values_from = c("IFN","IL1b","IL6","IL10","TNF"))
# write_csv(ck_wider, file = "/Users/arbones/Dropbox/SyncBriefcase/LAB/UK/APOE_switch/Georgia_bulk/20250917-cytokine_clean.csv")
```

### Within-group cytokine changes across conditions

For each genotype separately, cytokine concentrations are shown as medians with interquartile ranges. Pre- vs post-LPS plasma values and brain values were compared using non-parametric tests (p.overall).

```{r}
#| column: page

c1 <- compareGroups(prepost ~ IFN + IL1b + IL6 + IL10 + TNF, ck, method = 2)
t1 <- createTable(c1)

s1 <- strataTable(t1, "genotype")

export2md(s1, caption = "" )
```
As expected, there is a strong induction of all cytokines in plasma post-LPS and measurable levels in the brain. 
---

### Between-genotype comparisons

Cytokine values for +/+ and MC/+ are directly compared at each condition (pre, post, brain).

```{r}
#| column: page

c1 <- compareGroups(genotype ~ IFN + IL1b + IL6 + IL10 + TNF, ck, method = 2)
t1 <- createTable(c1)

s1 <- strataTable(t1, "prepost")

export2md(s1, caption = "" )
```

### Multiple-testing correction

To account for the 15 tests performed (5 cytokines × 3 conditions), the same Mann–Whitney genotype comparisons were re-computed and adjusted with the Benjamini–Hochberg procedure across all features.

```{r}
ck_padj <- ck |>
    mutate(condition = fct_recode(prepost,
                                  "Pre"  = "(Pre)",
                                  "Post" = "(Post)",
                                  "Brain" = "Brain")) |>
    pivot_longer(c(IFN, IL1b, IL6, IL10, TNF),
                 names_to = "cytokine", values_to = "value") |>
    group_by(condition, cytokine) |>
    summarise(p = wilcox.test(value ~ genotype, exact = FALSE)$p.value,
              .groups = "drop") |>
    mutate(p.BH = p.adjust(p, method = "BH")) |>
    pivot_wider(id_cols = cytokine,
                names_from = condition,
                values_from = c(p, p.BH))

kbl(ck_padj, digits = 3,
    caption = "Mann–Whitney p-values: raw and BH-adjusted across the 15 genotype × condition tests") |>
    kable_styling("striped", full_width = FALSE)
```

After Benjamini–Hochberg correction, no genotype comparison reaches significance, consistent with the manuscript's reported finding of no cytokine differences between 4s2 and 4s2M mice following LPS.
