---
title: "LPS Switch"
date: today
embed-resources: true
execute:
    echo: false
    warning: false
editor_options: 
  chunk_output_type: console
---
```{r}

library(tidyverse)
library(DESeq2)          # Framework for RNA-seq differential expression
library(patchwork)
library(kableExtra)
library(EnhancedVolcano)

# Read the raw gene count data from a tab-separated value (TSV) file.
counts <- read.table("~/Dropbox/SyncBriefcase/LAB/UK/APOE_switch/Georgia_bulk/nf-core_results/salmon_counts/salmon.merged.gene_counts.tsv", 
                       header = TRUE, sep = "\t") |>  # Specify that the file has a header and is tab-delimited.
  # Remove any duplicate rows based on the 'gene_name' column, keeping the first occurrence.
  distinct(gene_name, .keep_all = T) |> 
  # Set the 'gene_name' column as the row names of the data frame. DESeq2 requires gene identifiers as row names.
  column_to_rownames("gene_name") |> 
  # Remove the 'gene_id' and 'M2' columns, which are not needed for this analysis.
  select(-gene_id, -M2) |> 
  # Convert all remaining columns (which are sample counts) to integer format, as required by DESeq2.
  mutate(across(everything(), as.integer))


# Read the sample metadata from an Excel file into a data frame called 'md0'.
md0 <- readxl::read_xlsx("~/Dropbox/SyncBriefcase/LAB/UK/APOE_switch/Georgia_bulk/20250905_Bulk seq IDs.xlsx")

# Process the metadata.
md <- md0 |>
    # Set the 'altID' column as the row names for the metadata frame 'md'. 
    # These row names should correspond to the sample names (column names) in the 'counts' data frame.
    column_to_rownames("altID") 
    
# Reorder the rows of the metadata data frame ('md') to match the column order of the counts data frame ('counts').
# This is a CRITICAL step to ensure that the sample information in 'md' correctly aligns with the sample data in 'counts'.
md <- md[colnames(counts),] 

# identical(rownames(md),names(counts))
```

# Experiment summary 

Female microglia switch (MC+) and cre-negative (+/+) littermates were given tamoxifen to induce allelic switching at 11 months. 10 days later they were given intraperitoneal injections with LPS to cause inflammation. Brains were collected 24 hours after LPS injections.

```{r}

    
kbl(md) |> 
    kable_styling("striped")
```

# Bioinformatic methods

## RNAseq Data Processing and Analysis

Data processing was performed using the `nf-core/rnaseq` workflow (v3.19.0). The pipeline was executed with `Nextflow` v24.10.4 on the Morgan Computer Core high-performance computing cluster at the University of Kentucky. All analyses were containerized using Singularity.

**Quality Control and Preprocessing**: Raw sequencing reads underwent comprehensive quality assessment using `FastQC` to evaluate per-base sequence quality, GC content, sequence length distribution, and potential adapter contamination. Low-quality bases and adapter sequences were removed using `Trim Galore!`, which combines `Cutadapt` and `FastQC` functionalities to ensure optimal read quality for downstream analysis.

**Reference Genome and Annotation**: Trimmed reads were aligned to the mouse reference genome (Mus musculus GRCm39 primary assembly) obtained from Ensembl, using the corresponding GTF annotation file (Ensembl release 114) to define gene boundaries and features. 

**Read Alignment and Quantification**: `STAR` was used to perform splice-aware alignment of reads to the reference genome, generating coordinate-sorted BAM files. Simultaneously, `Salmon` quantified transcript abundance as raw counts and normalized as Transcripts Per Million (TPM) values. 


## Differential Expression Analysis (MC/+ _vs._ +/+)

Differential expression between experimental conditions was assessed using Wald tests implemented in `DESeq2` (version `r packageVersion("DESeq2")`) using the scaled gene count matrix generated by the `nf-core` pipeline. P-values were adjusted for multiple testing using the Benjamini-Hochberg false discovery rate (FDR) correction. Differentially expressed genes (DEGs) were identified using an adjusted p-value (FDR) < 0.05. 

Volcano plots were generated using the `EnhancedVolcano` R package (version `r packageVersion("EnhancedVolcano")`) to visualize the relationship between statistical significance (-log10 adjusted p-value) and biological effect size (log2 fold change) for all analyzed genes.Data quality was assessed through principal component analysis (PCA) to evaluate sample clustering and identify potential batch effects or outliers. 


---

# Results

Two filtering steps were applied to the raw gene counts:

1. genes with low expression were removed by retaining only those with a raw count of at least 10 in 5 or more samples.   
2. Following this, a second filter was applied to ensure robust expression, keeping only genes with a median expression count greater than 10 across all retained samples.  

```{r}
#| column: page
#| fig-width: 12

# Create an initial DESeqDataSet object from the raw count matrix and metadata.
# The design formula '~ genotype' tells DESeq2 to model gene expression based on the 'genotype' variable.
dds0 <- DESeqDataSetFromMatrix(countData = counts,
                               colData = md,
                               design = ~ genotype) 



# Define a filter to keep genes with sufficient expression.
# 'keep' will be a logical vector (TRUE/FALSE) for each gene.
# A gene is kept (TRUE) if it has 10 or more counts in at least 5 samples.
keep <- rowSums(counts(dds0) >= 10) >= 5

# Apply the filter to the DESeqDataSet object, creating a new object 'dds' with only the kept genes.
dds <- dds0[keep,]

# Set the reference level for the 'genotype' factor to "+/+".
# This ensures that all comparisons will be made against the "+/+" genotype.
dds$genotype <- relevel(dds$genotype, ref = "+/+")

# Calculate the median count for each gene across all samples.
median_expression <- rowMedians(counts(dds))

# Create a histogram to visualize the distribution of log-transformed median gene expression.
# This helps assess the overall expression levels in the dataset.

h <- ggplot(log(median_expression) |> as_tibble(), aes(value))+
    geom_histogram(fill = "grey80")+
    labs(x = "log10 (median expression)")+
    theme_minimal()+
    theme(axis.title = element_text(size=14),
          axis.text = element_text(size=12))


# Define a second filter to remove genes with very low median expression (less than or equal to 10).
keep2 <- median_expression > 10

# Apply the second filter to the 'dds' object.
dds <- dds[keep2,]

# Apply a Variance Stabilizing Transformation (VST) to the filtered count data.
# VST helps to make the variance more constant across the range of mean expression values, which is beneficial for PCA.
# 'blind = T' tells vst NOT to use the design formula (~genotype) when calculating the transformation.
dds_vst <- vst(dds, blind = T)

# Perform Principal Component Analysis (PCA) on the VST-transformed data.
# 'intgroup = "genotype"' specifies that points should be colored by genotype.
# 'returnData = T' ensures the function returns a data frame with PCA results for custom plotting.
pca <- plotPCA(dds_vst, intgroup = "genotype", returnData = T)

# Extract the percentage of variance explained by each principal component and round to the nearest whole number.
percentVar <- round(100 * attr(pca, "percentVar"))    

# Create a custom PCA plot using ggplot2.
p <- ggplot(pca, aes(PC1, PC2, color = genotype)) +
  geom_point(size = 3) +   # Plot points for each sample.
  # Label the x-axis with the percentage of variance explained by PC1.
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +  
  # Label the y-axis with the percentage of variance explained by PC2.
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +  
  # Ensure the scaling of the x and y axes is equal.
  coord_fixed() +    
  # Manually set the colors for the different genotypes.
  scale_color_manual("Genotype", values = c("#99c140","#e7b416")) +   
  # Use a minimal theme for a clean look.
  theme_minimal() +  
  # Adjust the font size of the axis titles.
  theme(axis.title = element_text(size =14))

# Use the 'patchwork' package's '|' operator to display the histogram (h) and the PCA plot (p) side-by-side.
h|p
```


```{r}
# Run the main DESeq2 differential expression analysis pipeline.
# This single function performs three steps:
# 1. estimateSizeFactors(): Calculates normalization factors.
# 2. estimateDispersions(): Estimates gene-wise dispersion.
# 3. nbinomWaldTest(): Fits a negative binomial model and performs a Wald test for each gene.
dds <- DESeq(dds)

# Extract the results table from the DESeq2 object.
# By default, this compares the last level of the factor against the reference level set earlier.
# The results are formatted as a DataFrame and then converted to a standard R data.frame for easier manipulation.
res <- results(dds, format = "DataFrame") |> 
    as.data.frame()

# This line is commented out, but if run, it would save the results data frame to a CSV file.
write.csv (res |> arrange(padj), file = "20250907-DEGs.csv")

# List the names of the coefficients that can be used for building contrasts with the results() function.
# This is useful for confirming the comparison being made or for specifying non-default comparisons.
# resultsNames(dds)
```


```{r}
# Create a volcano plot for differentially expressed genes (DEGs) in scWAT
EnhancedVolcano(res,
    # Label the points on the plot with gene names from the rownames of res_sc
    lab = rownames(res),
    # Set the main title of the plot
    title = "Differentially expressed genes (DEGs)",
    # Set the subtitle indicating the comparison groups
    subtitle = "MC/+ vs. +/+",
    ylab = bquote(~-Log[10] ~ italic(pvalue)),
    # Customize the legend labels for different categories of DEGs
    legendLabels = c("NS", 
                     expression(paste("|",Log[2], " FC","|", " >1")), 
                     "FDR < 0.05", 
                     expression(paste("FDR < 0.05 and ","|",Log[2], " FC","|", " >1"))),
    # Set the caption below the plot (empty in this case)
    caption = "",
    # Set the size and face of the labels on the plot
    labSize = 5, labFace = "italic",
    # Position the legend to the right of the plot
    legendPosition = 'no',
    # Draw lines connecting labels to their corresponding points
    drawConnectors = TRUE,
    # Connect labels to points in both directions (from left and right)
    directionConnectors = "both",
    # Allow for an infinite number of overlapping points to be labeled
    max.overlaps = Inf,
    # Set the fold change cutoff for highlighting DEGs
    FCcutoff = 0.001,
    # Specify the column to use for the p-value cutoff
    pCutoffCol = "pvalue",
    # Set the p-value cutoff for significance
    #pCutoff = 0.05,
    #cutoffLineType = "blank",
    # Set the x-axis to represent log2 fold change
    x = 'log2FoldChange',
    # Set the y-axis to represent the p-value
    y = 'pvalue',
    ylim = c(0,7),
    xlim = c(-3,2))
```

