---
title: "Gene Enrichment Analysis"
date: today
embed-resources: true
execute: 
  echo: false
  warning: false
  messages: false
format: html
---

```{r}
# fgsea: Fast Gene Set Enrichment Analysis
# https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html
library(fgsea)

# msigdbr: Access to MSigDB gene sets for multiple species
library(msigdbr)

# Data manipulation and visualization
library(tidyverse)
library(kableExtra)
```

### Gene Set Enrichment Analysis (GSEA)

Gene Set Enrichment Analysis (GSEA) was performed to identify biological pathways systematically altered by APOE2 overexpression in glial cells. This method evaluates whether genes within predefined gene sets are randomly distributed throughout a ranked gene list or concentrated at the extremes, indicating coordinated regulation of biological processes. All genes from the differential expression analysis were ranked by `log2 fold change` and tested against Gene Ontology Biological Process (GO:BP) gene sets from `MSigDB` for Mus musculus (minimum 30 genes, maximum 500 genes per pathway).

```{r}
#| fig-width: 10
#| fig-height: 7


#################################################################
### Mouse Gene Sets from MSigDB #################################
### https://www.gsea-msigdb.org/gsea/msigdb/mouse/genesets.jsp ##
#################################################################

set.seed(42)  # Set seed for reproducibility

# Load complete differential expression results (all genes, not just significant)
brain <- read_csv("/Users/arbones/Dropbox/SyncBriefcase/LAB/UK/APOE_switch/Georgia_bulk/analysis/20250909-DEGs.csv") |> 
    rename(gene = "...1")

# Create ranked gene list based on log2 fold change
# GSEA uses the ranking of ALL genes, not just significant ones
gene_ranks <- brain$log2FoldChange
names(gene_ranks) <- brain$gene
gene_ranks <- sort(gene_ranks, decreasing = TRUE)  # Sort high to low

# Retrieve Gene Ontology Biological Process gene sets from MSigDB
# MSigDB Collection M5 = ontology gene sets
# Subcollection BP = Biological Process
gene_sets <- msigdbr(
  species = "Mus musculus",      # Mouse
  db_species = "MM",              # Mouse species code
  collection = "M5",              # GO gene sets
  subcollection = "BP"            # Biological Process ontology
)

# Convert to list format required by fgsea
# Each element = one gene set (pathway) with its member genes
gene_sets_list <- split(gene_sets$gene_symbol, gene_sets$gs_name)

# Run Fast Gene Set Enrichment Analysis
# Tests whether genes in each pathway are randomly distributed in the ranked list
# or concentrated at the top (upregulated) or bottom (downregulated)
fgsea_results <- fgsea(
  pathways = gene_sets_list,      # Gene sets to test
  stats = gene_ranks,             # Ranked gene list
  eps = 0.0,                      # Precision for p-value estimation (0 = exact)
  minSize = 30,                   # Minimum genes per pathway
  maxSize = 500                   # Maximum genes per pathway
)



# Filter and format results
gsea <- fgsea_results |> 
  arrange(padj) |>                      # Sort by adjusted p-value
  filter(padj < 0.01) |>                # Keep only significant pathways
  arrange(desc(NES))                     # Sort by normalized enrichment score



# Display results table
# kbl(gsea |> select(pathway, NES, padj, size),
#  digits = 3) |> 
#  kable_styling(full_width = FALSE, "striped")
```

> GSEA of brain tissue revealed `r nrow(gsea)` significantly enriched pathways (>30 genes /pathway, FDR < 0.01)




#### Reducing Redundant Pathways with `collapsePathways`

To reduce redundancy arising from overlapping gene membership between related pathways, the `collapsePathways` function from the `fgsea` package was applied, consolidating the results to the most significantly enriched pathway from each cluster.

```{r}

# Subset gene_sets_list to only include significant pathways
gene_sets_significant <- gene_sets_list[gsea$pathway]


# CollapsePathways
# This identifies redundant pathways based on leading edge overlap
collapsed <- collapsePathways(
  fgseaRes = fgsea_results |> filter(padj<0.01),     # Your fgsea results (filtered)
  pathways = gene_sets_significant,     # Gene sets (filtered to significant)
  pval.threshold = 0.01,               # P-value threshold for significance
  stats = gene_ranks                    # Same ranked gene list used in fgsea
)

# Extract the main (representative) pathways
main_pathways <- collapsed$mainPathways

# Step 5: Filter your results to keep only main pathways
gsea_collapsed <- fgsea_results |> 
  filter(padj<0.01) |>
  filter(pathway %in% main_pathways) |>
  arrange(desc(NES)) |> 
  mutate(pretty_names = str_remove_all(pathway, "GOBP_"),
         pretty_names = str_replace_all(pretty_names, "_", " "))

# View how many pathways were collapsed
cat("Original significant pathways:", nrow(gsea), "\n")
cat("After collapsing:", nrow(gsea_collapsed), "\n")



```

Results are reported as Normalized Enrichment Score (NES, indicating direction and magnitude of enrichment), adjusted p-value (padj, statistical significance after FDR correction), and size (number of genes in the pathway present in the dataset).
```{r}
# Display results table
kbl(gsea_collapsed |> select(pretty_names, NES, padj, size),
  digits = 3) |>
  kable_styling(full_width = FALSE, "striped")
```

#### Key Enriched Pathways

The most prominently upregulated pathways were associated with protein synthesis and mitochondrial energy metabolism. Immune response pathways also showed significant upregulation. In contrast, pathways involved in neuronal function and synaptic regulation were predominantly downregulated. 
```{r}
#| fig-height: 10
ggplot(gsea_collapsed, aes(x = reorder(pretty_names, NES), y = NES)) +
  geom_col(aes(fill = NES > 0),alpha = 0.9) +
  coord_flip() +
  labs(
    title = "GSEA",
    x = "GO: Biological Process Pathways",
    y = "Normalized Enrichment Score (NES)") +
  theme_minimal() +
  theme(legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey98"))
```

```{r}
#| fig-height: 4
plotEnrichment(gene_sets_list[["GOBP_TRANSLATION_AT_SYNAPSE"]], 
               gene_ranks) +
  labs(title = "GO: BP Traslation at Synapse",
       subtitle = "NES = 3.12, padj <0.001")
```
_All overexpressed genes in this pathway are ribosomal_

```{r}
cat(gene_sets_list[["GOBP_TRANSLATION_AT_SYNAPSE"]], sep = ", ")
```
---

```{r}
#| fig-height: 4
plotEnrichment(gene_sets_list[["GOBP_PROTON_MOTIVE_FORCE_DRIVEN_ATP_SYNTHESIS"]], 
               gene_ranks) +
  labs(title = "GO:BP Proton Motive Force Driven ATP Synthesis",
       subtitle = "NES = 2.79, padj = <0.001")
```
_All overexpressed genes in this pathway are mitochondrial_

```{r}
cat(gene_sets_list[["GOBP_PROTON_MOTIVE_FORCE_DRIVEN_ATP_SYNTHESIS"]], sep = ", ")
```
---

```{r}
#| fig-height: 4
plotEnrichment(gene_sets_list[["GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN"]], 
               gene_ranks) +
  labs(title = "GO:BP Antigen Processing and presentation of peptide antigen",
       subtitle = "NES = 2.39, padj = <0.001")
```


```{r}
cat(gene_sets_list[["GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN"]], sep = ", ")
```
---

